Outline flowCore paper

### 2008-08-11 paper/manuscript.tex
changes to the manuscript based on suggestions by Robert
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33188 bc3139a8-67e5-0310-9ffc-ced21a209358

\documentclass[12pt]{article}
\usepackage{url}
\usepackage[authoryear,round]{natbib}
\usepackage{graphicx}

%%FIXME: I would use some markup for the package name(s)

@@

\section*{Introduction}
\subsection*{Computational tools for the analysis of FC-HCS data}
During the last several years, automation technologies have been
developed that enable the use of flow cytometry high content screening
(FC-HCS) to generate large, complex datasets in both basic and
clinical research applications \citep{brinkman2007hcf}. A serious
bottleneck in the interpretation of existing studies and the
application of FC-HCS to even larger, more complex problems is that
data management and data analysis methods have not advanced
sufficiently far from the methods developed for traditional
small-scale, tube-based studies.
%%FIXME: RG
%applications of flow cytometry (FCM) to small-scale, tube-based
%studies.

@@

In this paper, we describe a set of flexible and well structured
computational tools to efficiently analyze FC-HCS. This set of tools
is based on the open source statistical software R \citep{Rmain}, in
conjunction with the Bioconductor Project \citep{BIOC}. Our intent is
to provide a shared research platform that enables bioinformaticians,
computer scientists, and statisticians to work collaboratively with
biologists and clinicians to develop novel methods  
%%this reference is a bit odd here - you better explain what they say
%%and why you quote them 
\citep{lizard2007fca} for FCM data analysis. A critical component of
this approach is the development and implementation of standards that
will facilitate the adoption of these methods by both the larger
research community and commercial software vendors.  Consequently, we
are optimistic that this software platform will play a critical role
in addressing the bottlenecks described above that limit further
advances in the application of FC-HCS.

The computational tools we have developed are distributed in the R
software language as the Bioconductor package flowCore. flowCore is a
freely available, highly functional and extensible FCM data analysis
platform that enables researchers to efficiently handle FC-HCS data
and encourages open development of tools for their coherent
analysis. Among the issues addressed by flowCore are computationally
efficient data structures and a range of specialized methods for
compensation, transformation, and gating.  flowCore runs on Windows,
Mac OS X and Linux/Unix operating systems. Conceptually, the
underlying data structures that build the core of our software can be
implemented in any programming language, and we chose the R language
for its advantages in statistical programming and visualization. In
this paper, we describe the functionality and methods underlying
flowCore, and provide examples of their use.

\subsection*{R and Bioconductor}
R is a robust statistical programing environment which offers a wide
range of statistical and visualization methods developed for various
fields of application. In particular, the Bioconductor project
provides R software modules for biological and clinical data analysis.
It is widely used to process such data, spanning a diversity of
research fields including genomics, proteomics and cell biology
\citep{gentleman2006bos}. The R language and Bioconductor are
particularly geared towards the analysis of large complex datasets
that typically arise from high throughput experiments such as
microarray gene expression analysis or FC-HCS. Using flow cytometry as
a high-throughput technology requires suitable software infrastructure
to facilitate the rapid handling of the data. In our implementation of
flowCore we rely on two important lessons learned from the field of
gene expression data analysis: the first being the importance of data
structures that reflect the underlying data and facilitate the
manipulations that are of most interest, while the second is the
importance of a modular architecture that allows for many developers
to extend and use the underlying infrastructure and to combine tools
in complex work flows.

\subsection*{Existing data standards and conventions}
%%FIXME: I think that this section is mixing two ideas - the need for
%%data representation - eg the class hierarchy, and the analysis
%%tools. RG suggests starting by saying what they do -then point out
%%that for complex experiments there is data on the samples (time
%%points, drugs etc - we have examples to cite) and the need to
%%combine these.  The notion of classes from an OOP language provide
%%one coherent way to describe these richer data structures and
%%functions/methods can work on them.  We will describe the classes
%%and their implementation in R, but they could just as easily be
%%written in any other language (eg Java, C++), and if
%%programs/analysis tools written in those languages use similar
%%structures then communication is simplified, and users could more
%%easily move data between analysis tools.  I doubt that reflection is
%%needed here - but if so, it would allow a naive function to get
%%information about the data structure (it is hard for it to use that
%%information though)

Currently, data from FCM experiments are stored in single files
according to the Flow Cytometry Standard (FCS) \citep{seamer1997pnd}.
However, recent developments in high-throughput FCM are shifting the
focus of interest away from single-tube based measurements towards
large and complex experimental designs with dozens of covariates and
influencing factors. Self-contained reflective data
%%FIXME: might need to explain reflective - or at least how you are
%%going to use this - we may need a graphic that shows the flow data,
%%data about the dyes antibodies but also about the samples etc - and
%%the need to keep this coordinated through an analysis
structures are a prerequisite to allow for coordinated analysis of
such experiments.  Many of the currently available software solutions
offer only limited support for such structures, or make use of
containers like XML or binary storage formats that are designed
specifically for particular user interfaces and hence are not easily
amenable to programmatical access. In addition, the closed-source
nature of these products often makes it impractical to integrate them into
analysis pipelines.

Traditionally, the majority of FCM experiments have been analyzed by
manual data inspection in one or two dimensions, or by very basic
comparisons of summary statistics. We believe that these approaches,
in addition to being extremely expensive and labor intensive, do not
fully address the highly complex nature of FCM data; in particular,
they discard many of the fundamental aspects of the data, such as its
underlying distribution and its high-dimensional nature. Furthermore,
the subjective character of manual analyses are a major obstacle to
reproducibility. For FC-HCS data, unassisted manual inspection is
extremely time consuming, and robust statistical methods need to be
developed to point investigators to interesting aspects of the data,
or to potential problems. While the expert knowledge of immunologists
and researchers remains crucial for the understanding of FCM data, we
believe that collaboration with other research fields such as
statistics and computer science can greatly improve the relevance of
FCM in today's high-throughput paradigm.

These issues are currently being addressed by emerging new flow
cytometry data standards developed in collaboration with the ISAC Data
Standards Task Force.  For example, the Gating-ML Candidate
Recommendation (CR) \textit{Fixme Ryan or Joseph: How do we cite
  this?}  represents a proposal on how to form unambiguous XML-based
gate definitions, which can facilitate the interchange and validation
of data between different software packages. Gates may be ordered into
a hierarchical structure to describe a gating strategy. This enables
the encoding of a simple analytical work flow, in a way that allows us
to reconstruct the analysis programmatically.

The flowCore framework presented here can import raw data FCS files
along with their complete set of file-specific meta information
(Figure~\ref{fig1:FrameWork}).  Moreover, it is a software
implementation of the Gating-ML CR, which makes it possibile to
integrate flowCore in existing work flows and to communicate with
any other FCM tool that adheres to the proposed standard. Adherence
to standards also plays a critical role in the ability of new methods
based on flowCore to find there way back into the standard practices
for FCM data analysis.  Note that the data structures defined in
flowCore aim to model large sets of FCM data in a natural and coherent
way; they are not specific to the R language and can be implemented in
any other programming language. Ideally, different software programs
would communicate on the level of these data structures rather than a
high-level markup language.

%%FIXME: class hierarchy is probably not going to be well understood
%%by readers - you might need to work that into a more coherent
%%general discussion
flowCore is not a GUI-driven software, and all operations are done
using a command line interface.  It is possible to add a more
elaborate user interface on top of this infrastructure, however the
focus in this paper is on a programmatic approach to enable the
convenient development of novel analysis methods and automation of
complex analysis approaches.  By taking the burden of data management
from the programmer, and by providing well-defined programming
interfaces (APIs) and a structured class hierarchy, it is possible to
readily test new ideas and to easily extend the framework's
functionality.

@@

FlowCore's primary task is the representation and basic manipulation
of FCM data. This is accomplished through a data model very close to
that adopted by other Bioconductor packages. All information from a
single FCS file, i.e., the collection of events and the accompanying
meta data, is stored in one single container. We call the structure
that hold this data a flowFrame. Raw data values as well as associated
metadata of a flowFrame can be accessed programmatically. Most
commonly, the metadata consist of descriptors of the stains used in
the experiment and the respective measurement channels, information
about compensation performed at the instrument side and any additional
keywords the user deems to be important to annotate the data. A number
of quality checks are performed during the creation of a flowFrame to
ensure data integrity.

%%again some markup for the class names would help the reader
%% stick with annotation or metadata not both
\subsubsection*{flowSet}
In high-throughput FCM, many of the analysis tasks need to be
performed consistently across multiple flow measurements, hence we
introduce the concept of a collection of flowFrames called a
flowSet. A flowSet is a container for multiple flowFrames along with
relevant information associated with each individual frame such as
descriptions of the cell sample, the treatment to which the sample was
subjected, or the location of that sample in a 96 or 384 well plate.
FlowSets manage the consistent application of operations on the
individual flowFrames and shift the burden of keeping score of the
annotation from the user to the infrastructure, thus reducing the risk
of errors (\textit{e.g.}, mixups of sample labels). The flowSet
structure can be readily extended to incorporate the potentially
complex annotation associated with even larger FC-HCS experiments such
as clinical trials where hundreds patients might provide samples at
different points in time over the course of the experiment.  Access to
flowCore data structures and methods has already enabled the
development of new Bioconductor packages, such as plateCore, that
facilitate the analysis of plate-based FC-HCS data and make handling
the annotation and processing the data easier.

%%Figure1 Diagram of the data structures and the basic operations
\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{Figure1-flowCoreFrameWork.jpg}
\caption{\label{fig1:FrameWork}{For each experiment, FCS files (as
    individual flowFrames), phenotypic and meta data are stored in a
    flowSet. Each flowFrame in a flowSet corresponds to one FCS
    file. All basic operations (e.g., compensation, transformation,
    gating) can be applied to either single flowFrames or a flowSet
    simultaneously.}}
\end{figure}

%%FIXME: if possible I would make the box that has Sample Meta-data in
%%it, be more of a table view (Age Sex Time with numbers and then it
%%might be more obvious what is happening

@@

The basic operations in FCM analyses are typically the same: the data
need to be compensated (if that was not already done on the instrument)
and transformed, and sup-populations of interest need to be selected
based on a set of (predominantly sequential) gates. All software
solutions for FCM analysis offer support for
these operations, typically in an interactive, GUI-driven interface. The
approach we have taken in flowCore is to abstractly describe these
operations and build a set of tools to perform them on both flowFrames
and flowSets. While transformation, and to a certain extent
compensation, are fairly routine operations with only limited
potential for improvement, being able to implement new methodologies
for gating of FCM data, and extending the capabilities of flowCore
through object oriented programming are features that clearly sets our
framework apart from other FCM analysis tools. By factoring out as
much of the bookkeeping as possible, programmers can focus on the
actual operations rather then having to deal with the tedious details
of data integration and access. Third-party methods can act on their
own as first-class citizens in the analysis framework, without
breaking the work flow or the basic infrastructure. This design allows
for the straightforward extension of flowCore's capabilities, and has
already fostered the development of a number of valuable add-ons
\citep{lo2008agf,sarkar2008ufv} \textit{(Fixme Errol: Will plateCore
  be in Bioconductor soon. It should be if we want to cite it.)}.

@@

Data transformation is essential for both data visualization and
modelling \citep{lo2008agf}. The major transformations that are
routinely used in FCM analysis have been implemented in flowCore
(\textit{e.g.,} quadratic, log or arcsinh, see Table~\ref{table1} for a
complete list). Furthermore, the design of the R language makes it
easy to define arbitrary functions to apply to the data of individual
flowFrames or entire flowSets, respectively. Compensation is
available for both flowFrames and flowSets, and the software also
offers functionality to compute spillover or compensation matrices
from a set of appropriate compensation samples.

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l|l|}
\hline
\multicolumn{2}{|c|}{Data Transformations} \\
\hline
linear & $ax + b$ \\
quadratic & $ax^2 + bx + c$ \\
natural logarithm & $ln(x)(r/d)$ \\
logarithm & $log(x)(r/d)$ \\
biexponential & need equation \\
logicle& need equation \\
truncate & $x_{x<=a} = a$ \\
scale & $(x-a)/(b-a)$ \\
split-scale & combination of linear and logarithm \\
arcsinh & $asinh(a + bx)+c$ \\
\hline
\end{tabular}
\caption{\label{table1}Data transformations implemented in flowCore.}
\end{center}
\end{table}

@@

In flowCore, gating operations are represented by classes that can be
extended in an object-oriented manner (Table~\ref{table2}). Basic gate
types such as rectangular gates, ellipses and polygon gates are
implemented as part of the framework. In addition, we introduce the
notion of data-driven gates, or filters, for which the necessary
parameters are computed based on the properties of the underlying
data, for instance by modelling data distribution or by density
estimation. The ability to programmatically access gates is a
prerequisite for automated or semi-automated gating. By utilizing a
unified interface for all different types of gates, the user is able
to subset data sets as well as to create summary statistics, for
example the proportions of events falling in a single gate or in a
combination of gates. Complex combinations and hierarchies of gates
can be captured in objects of class filterSet, allowing to apply
multi-step gating strategies. The definition of gates in flowCore
follows the Gating-ML CR, thus any flowCore gating strategy can be
reproduced by any other software that also adheres to the standard.

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l|l|}
\hline
\multicolumn{2}{|c|}{Gates} \\
\hline
rectangleGate & n-dimensional rectangular regions \\
quadGate & quadrant regions in two dimensions \\
polygonGate & polygonal regions in two dimensions \\
polytopeGate & generalization of polygon in n dimensions \\
ellipsoidGate & n-dimensional ellipsoid region \\
\hline
\multicolumn{2}{|c|}{Filters} \\
\hline
sampleFilter & random sub-sampling\\
expressionFilter & results of a boolean expression \\
kmeansFilter & K-means clustering \\
norm2Filter & bivariate normal distribution \\
curv1Filter & local density regions in 1D \\
curv2Filter & density regions in 2D \\
timeFilter & abnormal data acquisition over time \\
\hline
filterSet & gating strategies \\
\hline
\end{tabular}
\caption{\label{table2}Filter and gate classes implemented in flowCore.}
\end{center}
\end{table}

@@

In addition to the flowCore package that offers basic infrastructure,
we have implemented a range of additional Bioconductor packages that
are dedicated to more specific tasks of FCM data analysis. The flowViz
package \citep{sarkar2008ufv} provides sophisticated data
visualization tools, that make use of multivariate trellis plotting
\citep{FIXME:citeDeepayansBook}.  These functions can be used to
quickly generate customized plots for extended cytometry data sets for
both direct data inspection and quality control.  The objects
annotation information can be used to arrange the layout and
composition of the plots.  Furthermore, the design and the API of the
visualization software is very generic, and users can readily extend
its capabilities by providing self-defined plotting functions.  The
flowQ package offers more advanced quality assurance (QA) methodology
and a framework to create interactive web-based reports of QA results.
Most of flowUtil's content deals with data import and export using the
flow-cytometry specific standard markup languages (FIXME:it implements
the standards doesn't it?).  Finally, flowStats provides more
elaborate statistical methods that are relevant in the context of flow
cytometry data analysis.

\section*{Results}
%%FIXME: we seem to mention other packages a lot - some reduction and
%%cohesion is going to be needed

The flowCore package has been successfully applied in the analysis of
several datasets \citep{gasparetto2004ice,brinkman2007hcf}, and
several other Bioconductor packages are making use of its code
base. As an example, \cite{lo2008agf} have recently developed an
automatic gating approach via robust model-based clustering using
flowCore's data model and infrastructure and implemented this in the
Bioconductor package flowClust. Another package, plateCore, providing
more specialized support for experiments conducted on microtitre
plates is under development.

A complete documentation of the flowCore software is beyond the scope
of this publication.  Much more comprehensive documentation and users
guide information with programmatic examples is available online
(\url{http://bioconductor.org/packages/2.2/bioc/html/flowCore.html})
and as part of the package distribution. Here, we want to briefly
exemplify some of the software's key features, that is, the coherent
treatment of all samples in a potentially large experiment, the
concept of data-driven automated gating, the integration of existing
software into the framework, and the generation of publication-quality
graphics for data visualization.

Data analysis for most experiments usually begins with a quality
assurance step, and we can use functionality from the flowQ package to
create an HTML report that highlights potential quality
issues. Assuming that the data has already been imported as the
flowSet object ``dat'', the following simple line of code produces the
quite complex output:

@@

The report is interactive and provides drill-down to more detailed
aspects of the analysis, starting from a concise overview
(Figure~\ref{flowQ}). The design of flowCore's data model allows for a
coherent treatment of all the samples, hence we are able to compare
features between individuals, or between groups of individuals, based
on the available annotation information.

@@

Static gating for all samples in a high-throughput FCM experiment is
often not an option, since the measured variables tend to vary between
different treatments, over time or between different experiment
batches. Automated or data-driven gating tries to estimate the gating
regions from the underlying data, thus providing a fast objective
solution to the analysis of potentially very large and diverse data
sets \cite{lo2008agf}. One of the automated gating methods implemented
in flowCore is based on identifying areas of significant curvature in
a kernel density estimate of the data \citep{wand2008}. Assuming that
the regions of interest are of high density, the software is able to
reliably detect them in a one or two-dimensional density landscape.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=0.75\textwidth]{xyplot}
\caption{\label{xyplot}%
Scatterplot matrix of a single flowSet with outlines of the gating
regions identified by an automated gating operation.}
\end{figure}

>>>>

\documentclass[12pt]{article}
\usepackage{url}
\usepackage[authoryear,round]{natbib}
\usepackage{graphicx} 
\newcommand{\Rpackage}[1]{{\textsf{#1}}}
\newcommand{\Rclass}[1]{{\textit{#1}}}

@@

\section*{Abstract}
\subsubsection*{Background}
Recent advances in automation technologies have enabled the use of
flow-cytometry high content screening (FH-HCS). However, data
management and data analysis methods have not advanced sufficiently
far from the inital small-scale studies to support the large complex
data sets generated in FH-HCS applications.
\subsubsection*{Methods}
In this manuscript, we present \Rpackage{flowCore} a set of flexible
and well structured computational tools for FC-HCS data analysis based
on the open source statistical software R in conjunction with the
Bioconductor Project.
\subsubsection*{Results}
Through \Rpackage{flowCore}, we provide the FCM community a shared
research platform that enables bioinformaticians, computer scientists,
and statisticians to work collaboratively with biologists and
clinicians to develop novel methods for FCM data analysis.
\subsubsection*{Conclusions}
We hope that our framework will be the foundation for fruitful shared
research by many collaborators from multiple scientific fields and
will help resolve bottlenecks that currently prevent the further
development of and deployment of FC-HCS to increasingly complex and
important scientific and clinical applications.

\section*{Introduction}
\subsection*{Computational tools for the analysis of FC-HCS data}
During the last several years, automation technologies have been
developed that enable the use of flow cytometry high content screening
(FC-HCS) to generate large, complex datasets in both basic and
clinical research applications \citep{brinkman2007hcf}. A serious
bottleneck in the interpretation of existing studies and the
application of FC-HCS to even larger, more complex problems is that
data management and data analysis methods have not advanced
sufficiently far from the methods developed for applications of flow
cytometry (FCM) to small-scale, tube-based studies.

@@

In this paper, we describe a set of flexible and well structured
computational tools to efficiently analyze FC-HCS. This set of tools
is based on the open source statistical software R \citep{Rmain}, in
conjunction with the Bioconductor Project \citep{BIOC}. Our intent is
to provide a shared research platform that enables bioinformaticians,
computer scientists, and statisticians to work collaboratively with
biologists and clinicians to develop novel methods for FCM data
analysis, a process deemed crucial by many for the further development
of the technology \citep{lizard2007fca}. A critical component of this
approach is the development and implementation of standards that will
facilitate the adoption of these methods by both the larger research
community and commercial software vendors.  Consequently, we are
optimistic that this software platform will play a critical role in
addressing the bottlenecks described above that limit further advances
in the application of FC-HCS.

The computational tools we have developed are distributed in the R
software language as the Bioconductor package
\Rpackage{flowCore}. \Rpackage{flowCore} is a freely available, highly
functional and extensible FCM data analysis platform that enables
researchers to efficiently handle FC-HCS data and encourages open
development of tools for their coherent analysis. Among the issues
addressed by \Rpackage{flowCore} are computationally efficient data
structures and a range of specialized methods for compensation,
transformation, and gating.  \Rpackage{flowCore} runs on Windows, Mac
OS X and Linux/Unix operating systems. Conceptually, the underlying
data structures that build the core of our software can be implemented
in any programming language, and we chose the R language for its
advantages in statistical programming and visualization. In this
paper, we describe the functionality and methods underlying
\Rpackage{flowCore}, and provide examples of their use.

\subsection*{R and Bioconductor}
R is a robust statistical programing environment which offers a wide
range of statistical and visualization methods developed for various
fields of application. In particular, the Bioconductor project
provides R software modules for biological and clinical data analysis.
It is widely used to process such data, spanning a diversity of
research fields including genomics, proteomics and cell biology
\citep{BIOC}. The R language and Bioconductor are
particularly geared towards the analysis of large complex datasets
that typically arise from high throughput experiments such as
microarray gene expression analysis or FC-HCS. Using flow cytometry as
a high-throughput technology requires suitable software infrastructure
to facilitate the rapid handling of the data. In our implementation of
\Rpackage{flowCore} we rely on two important lessons learned from the field of
gene expression data analysis: the first being the importance of data
structures that reflect the underlying data and facilitate the
manipulations that are of most interest, while the second is the
importance of a modular architecture that allows for many developers
to extend and use the underlying infrastructure and to combine tools
in complex work flows.

\subsection*{Existing data standards and conventions}
Traditionally, the majority of FCM experiments have been analyzed by
manual data inspection in one or two dimensions, or by very basic
comparisons of summary statistics and most of the currently available
analysis tools are designed to reflect this workflow.  We believe that
these approaches, in addition to being expensive and labor intensive,
do not fully address the highly complex nature of FCM data; in
particular, they disregard many of the fundamental aspects of the
data, such as its underlying distribution and its high-dimensional
nature. Furthermore, the subjective character of manual analyses are a
major obstacle to reproducibility. For FC-HCS data, unassisted manual
inspection is extremely time consuming, and robust statistical methods
need to be developed to point investigators to interesting aspects of
the data, or to potential problems. While the expert knowledge of
immunologists and researchers remains crucial for the understanding of
FCM data, we believe that collaboration with other research fields
such as statistics and computer science can greatly improve the
relevance of FCM in today's high-throughput paradigm.

Currently, data from FCM experiments are stored in single files
according to the Flow Cytometry Standard (FCS) \citep{seamer1997pnd}.
However, recent developments in high-throughput FCM are shifting the
focus of interest away from single-tube based measurements towards
large and complex experimental designs with dozens of covariates and
influencing factors. For example, experiments consist of large numbers
of samples from different patients, measured at different timepoints
\citep{brinkman2007hcf} or following different drug treatments
\citep{gasparetto2004ice}. Modern FCM data analysis tools have to deal
with an additional layer of sample metadata and they need to provide
infrastructure to process and to compare groups of samples in a
concise and coordinated manner.

The notion of classes from an object-oriented programming language
provides one coherent way to describe these richer data structures,
and functions or methods that work on them allow for interaction and
manipulation. Many of the currently available software solutions offer
only limited support for such self-contained structures, or make use
of containers like XML or binary storage formats that are designed
specifically for particular user interfaces and hence are not easily
amenable to programmatical access. In addition, the closed-source
nature of these products often makes it impractical to integrate them
into analysis pipelines. In this manuscript we describe classes for
FCM data analysis and their implementation in R, however, they could
just as easily be implemented in any other language (eg Java,
C++). Software written in those languages could use similar data
structures, thereby simplifying communication and the interchange of
data between analysis tools.

The \Rpackage{flowCore} framework presented here can import and
process raw data FCS files along with their complete set of
file-specific metadata (Figure~\ref{fig1:FrameWork}).  Moreover, it is
a software implementation of the Gating-ML CR, an emerging new flow
cytometry data standard developed in collaboration with the ISAC Data
Standards Task Force, which makes it possibile to integrate
\Rpackage{flowCore} in existing work flows and to communicate with any
other FCM tool that adheres to the proposed standard. Adherence to
standards also plays a critical role in the ability of new methods
based on \Rpackage{flowCore} to find their way back into the standard
practices for FCM data analysis. \Rpackage{flowCore} is not a
GUI-driven software, and all operations are done using a command line
interface.  It is possible to add a more elaborate user interface on
top of this infrastructure, however the focus in this paper is on a
programmatic approach to enable the convenient development of novel
analysis methods and automation of complex analysis approaches.  By
taking the burden of data management from the programmer, and by
providing well-defined programming interfaces (APIs), it is possible
to readily test new ideas and to easily extend the framework's
functionality.

@@

\Rpackage{flowCore}'s primary task is the representation and basic
manipulation of FCM data. This is accomplished through a data model
very close to that adopted by other Bioconductor packages. All
information from a single FCS file, i.e., the collection of events and
the accompanying metadata, is stored in one single container. We call
the structure that hold this data a \Rclass{flowFrame}. Raw data
values as well as associated metadata of a \Rclass{flowFrame} can be
accessed programmatically. Most commonly, the metadata consist of
descriptors of the stains used in the experiment and the respective
measurement channels, information about compensation performed at the
instrument side and any additional keywords the user deems to be
important to annotate the data. A number of quality checks are
performed during the creation of a \Rclass{flowFrame} to ensure data
integrity.

\subsubsection*{flowSet}
In high-throughput FCM, many of the analysis tasks need to be
performed consistently across multiple flow measurements, hence we
introduce the concept of a collection of \Rclass{flowFrames} called a
\Rclass{flowSet}. A \Rclass{flowSet} is a container for multiple
\Rclass{flowFrames} along with relevant information associated with
each individual frame such as descriptions of the cell sample, the
treatment to which the sample was subjected, or the location of that
sample in a 96 or 384 well plate.  \Rclass{flowSets} manage the
consistent application of operations on the individual
\Rclass{flowFrames} and shift the burden of keeping score of the
metadata from the user to the infrastructure, thus reducing the risk
of errors (\textit{e.g.}, mixups of sample labels). The
\Rclass{flowSet} structure can be readily extended to incorporate the
potentially complex metadata associated with even larger FC-HCS
experiments such as clinical trials where hundreds patients might
provide samples at different points in time over the course of the
experiment.

%%Figure1 Diagram of the data structures and the basic operations
\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{Figure1-flowCoreFrameWork.jpg}
\caption{\label{fig1:FrameWork}{For each experiment, FCS files (as
    individual \Rclass{flowFrames}), phenotypic and metadata are
    stored in a \Rclass{flowSet}. Each \Rclass{flowFrame} in a
    \Rclass{flowSet} corresponds to one FCS file. All basic operations
    (e.g., compensation, transformation, gating) can be applied to
    either single \Rclass{flowFrames} or a \Rclass{flowSet}
    simultaneously.}}
\end{figure}

@@

The basic operations in FCM analyses are typically the same: the data
need to be compensated (if that was not already done on the
instrument) and transformed, and sup-populations of interest need to
be selected based on a set of (predominantly sequential) gates. All
software solutions for FCM analysis offer support for these
operations, typically in an interactive, GUI-driven interface. The
approach we have taken in \Rpackage{flowCore} is to abstractly
describe these operations and build a set of tools to perform them on
both \Rclass{flowFrames} and \Rclass{flowSets}. While transformation,
and to a certain extent compensation, are fairly routine operations
with only limited potential for improvement, being able to implement
new methodologies for gating of FCM data, and extending the
capabilities of \Rpackage{flowCore} through object oriented
programming are features that clearly sets our framework apart from
other FCM analysis tools. By factoring out as much of the bookkeeping
as possible, programmers can focus on the actual operations rather
then having to deal with the tedious details of data integration and
access. Third-party methods can act on their own as first-class
citizens in the analysis framework, without breaking the work flow or
the basic infrastructure. This design allows for the straightforward
extension of \Rpackage{flowCore}'s capabilities, and has already
fostered the development of a number of valuable add-ons
\citep{lo2008agf,sarkar2008ufv} \textit{(Fixme Errol: Will plateCore
  be in Bioconductor soon. It should be if we want to cite it.)}.

@@

Data transformation is essential for both data visualization and
modelling \citep{lo2008agf}. The major transformations that are
routinely used in FCM analysis have been implemented in
\Rpackage{flowCore} (\textit{e.g.,} quadratic, log or arcsinh, see
Table~\ref{table1} for a complete list). Furthermore, the design of
the R language makes it easy to define arbitrary functions to apply to
the data of individual \Rclass{flowFrames} or entire
\Rclass{flowSets}, respectively. Compensation is available for both
\Rclass{flowFrames} and \Rclass{flowSets}, and the software also
offers functionality to compute spillover or compensation matrices
from a set of appropriate compensation samples.

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l|l|}
\hline
\multicolumn{2}{|c|}{Data Transformations} \\
\hline
linear & $ax + b$ \\
quadratic & $ax^2 + bx + c$ \\
natural logarithm & $ln(x)(r/d)$ \\
logarithm & $log(x)(r/d)$ \\
biexponential & $root(a*e^{(b*y)}-c*e^{(-d+y)}+f)$ \\
logicle& $root(Te^{-(m-w)}*(e^{(y-w)}-p^2e^{-(y-w)/p}+p^2-1)$ \\
truncate & $x_{x<=a} = a$ \\
scale & $(x-a)/(b-a)$ \\
split-scale & combination of linear and logarithm \\
arcsinh & $asinh(a + bx)+c$ \\
\hline
\end{tabular}
\caption{\label{table1}Data transformations implemented in
  \Rpackage{flowCore}.}
\end{center}
\end{table}

@@

In \Rpackage{flowCore}, gating operations are represented by classes
that can be extended in an object-oriented manner
(Table~\ref{table2}). Basic gate types such as rectangular gates,
ellipses and polygon gates are implemented as part of the
framework. In addition, we introduce the notion of data-driven gates,
or filters, for which the necessary parameters are computed based on
the properties of the underlying data, for instance by modelling data
distribution or by density estimation. This approach is fundamentally
different from the traditional application of static gating regions
across samples, as it is able to take into accounts unforeseen changes
in signal intensities. \textit{e.g.} drifts in the instrumentation
over time or sample variablity.

The ability to programmatically access gates is a prerequisite for
automated or semi-automated gating. By utilizing a unified interface
for all different types of gates, the user is able to subset data sets
as well as to create summary statistics, for example the proportions
of events falling in a single gate or in a combination of
gates. Complex combinations and hierarchies of gates can be captured
in objects of class \Rclass{filterSet}, allowing to apply multi-step
gating strategies. The definition of gates in \Rpackage{flowCore}
follows the Gating-ML CR, thus any \Rpackage{flowCore} gating strategy
can be reproduced by any other software that also adheres to the
standard.

Gating, as well as all other operations in flowCore, can be applied
over each individual frame in a \Rclass{flowSet}, and summary methods
provide information about the outcome of these operations. In
addition, the result of a gating operation can be used to subset the
input \Rclass{flowFrame} or \Rclass{flowSet}, either by filtering out
negative events or by splitting in multiple subpopulations. This
design allows to easily combine all of \Rpackage{flowCore}'s
components into complex work flows.

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l|l|}
\hline
\multicolumn{2}{|c|}{Gates} \\
\hline
rectangleGate & n-dimensional rectangular regions \\
quadGate & quadrant regions in two dimensions \\
polygonGate & polygonal regions in two dimensions \\
polytopeGate & generalization of polygon in n dimensions \\
ellipsoidGate & n-dimensional ellipsoid region \\
\hline
\multicolumn{2}{|c|}{Filters} \\
\hline
sampleFilter & random sub-sampling\\
expressionFilter & results of a boolean expression \\
kmeansFilter & K-means clustering \\
norm2Filter & bivariate normal distribution \\
curv1Filter & local density regions in 1D \\
curv2Filter & density regions in 2D \\
timeFilter & abnormal data acquisition over time \\
\hline
filterSet & gating strategies \\
\hline
\end{tabular}
\caption{\label{table2}Filter and gate classes implemented in
  \Rpackage{flowCore}.}
\end{center}
\end{table}

@@

In addition to the \Rpackage{flowCore} package that offers basic
infrastructure, we have implemented a range of additional Bioconductor
packages that are dedicated to more specific tasks of FCM data
analysis. The \Rpackage{flowViz} package \citep{sarkar2008ufv}
provides sophisticated data visualization tools, that make use of
multivariate trellis plotting \citep{lattice}.  These functions can be
used to quickly generate customized plots for extended cytometry data
sets for both direct data inspection and quality control.  The objects
metadata information can be used to arrange the layout and composition
of the plots.  Furthermore, the design and the API of the
visualization software is very generic, and users can readily extend
its capabilities by providing self-defined plotting functions.  The
\Rpackage{flowQ} package offers more advanced quality assurance (QA)
methodology and a framework to create interactive web-based reports of
QA results.  Most of \Rpackage{flowUtil}'s content deals with data
import and export, implementing the flow-cytometry specific standard
markup language.  Finally, \Rpackage{flowStats} provides more
elaborate statistical methods that are relevant in the context of flow
cytometry data analysis. \cite{lo2008agf} have recently developed an
automatic gating approach via robust model-based clustering using
\Rpackage{flowCore}'s data model and infrastructure which is
implemented in the Bioconductor package \Rpackage{flowClust} Another
package, \Rpackage{plateCore}, providing more specialized support for
experiments conducted on microtitre plates and facilitiating the
handling of spatial metadata, is under development.

\section*{Results}
The \Rpackage{flowCore} package has been successfully applied in the
analysis of several datasets
\citep{gasparetto2004ice,brinkman2007hcf}. A complete documentation of
the \Rpackage{flowCore} software is beyond the scope of this
publication.  Much more comprehensive documentation and users guide
information with programmatic examples is available online
(\url{http://bioconductor.org/packages/2.2/bioc/html/flowCore.html})
and as part of the package distribution. Here, we want to briefly
exemplify some of the software's key features, that is, the coherent
treatment of all samples in a potentially large experiment, the
concept of data-driven automated gating, the integration of existing
software into the framework, and the generation of publication-quality
graphics for data visualization.

Data analysis for most experiments usually begins with a quality
assurance step, and we can use functionality from the \Rpackage{flowQ}
package to create an HTML report that highlights potential quality
issues. Assuming that the data has already been imported as the
\Rclass{flowSet} object ``dat'', the following simple line of code
produces the quite complex output:

@@

The report is interactive and provides drill-down to more detailed
aspects of the analysis, starting from a concise overview
(Figure~\ref{flowQ}). The design of \Rpackage{flowCore}'s data model
allows for a coherent treatment of all the samples, hence we are able
to compare features between individuals, or between groups of
individuals, based on the available metadata information.

@@

Static gating for all samples in a high-throughput FCM experiment is
often not an option, since the measured variables tend to vary between
different treatments, over time or between different experiment
batches. Automated or data-driven gating tries to estimate the gating
regions from the underlying data, thus providing a fast objective
solution to the analysis of potentially very large and diverse data
sets \cite{lo2008agf}. One of the automated gating methods implemented
in \Rpackage{flowCore} is based on identifying areas of significant
curvature in a kernel density estimate of the data
\citep{wand2008}. Assuming that the regions of interest are of high
density, the software is able to reliably detect them in a one or
two-dimensional density landscape.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=0.75\textwidth]{xyplot}
\caption{\label{xyplot}%
  Scatterplot matrix of a single \Rclass{flowSet} with outlines of the
  gating regions identified by an automated gating operation.}
\end{figure}
